\documentclass[a4paper]{article}
\begin{document}
\title{gdbank annotation guidelines v. 0.10}
\author{Colin Batchelor}
\maketitle

\tableofcontents

\section{Background}

These guidelines assume at least a basic working knowledge of written Scottish Gaelic, particularly how to identify nouns, verbs, adjectives and determiners.

\subsection{General instructions}

\noindent {\bf 1.} You may use a dictionary to elucidate meaning.

\noindent {\bf 2.} You may also refer to the English gloss if such a thing has been provided, for example in {\it An Litir Bheag}.

\noindent {\bf 3.} Gaelic is VSO. As a result it is very likely that the first word you come across will be either a verb or a preverbal particle.
The general pattern for preverbal particles is $\oslash$ (declarative), \textit{cha/chan} (negative), \textit{a/am/an} (interrogative), \textit{nach} (negative interrogative).

\noindent{\bf 68.} If you get absolutely stuck as to the categorial annotation, put X as the category.

\noindent {\bf 77.} For the categorial annotation, if a word has been missed out, possibly for scansion or just in error, then don't modify the types of words that would otherwise depend on it.

\noindent {\bf 82.} In the categorial annotation, annotate the tokens bearing in mind the type-changing rules in section~\ref{section:typechanging}. For example, preposition types will include \texttt{PP} even though in many cases they will be type-changed to \texttt{N\textbackslash N} or \texttt{S\textbackslash S}. Note down any type-changing rules that are needed but missing.

\noindent {\bf 97.} Out of scope for annotation: number, gender.
***Case is not explicitly annotated but this may well be wrong.

\subsection{Format}

\noindent {\bf 5.} gdbank follows the CoNLL-X format.
You \textbf{must} fill in the following columns: (1) token number, (2) surface, (3) lemma, (4) CPOSTAG (with the Universal POS tag detailed below), (5) POSTAG (to match CPOSTAG, for the moment), (6) FEATS, (7) head and (8) dependency type.
You \textbf{may} specify the others; if not put an underscore in them.

\section{Tokenization}

\newcommand{\ADV}{\texttt{ADV}}
\newcommand{\NOUN}{\texttt{NOUN}}
\newcommand{\PRT}{\texttt{PRT}}
\newcommand{\ROOT}{\texttt{ROOT}}
\newcommand{\VERB}{\texttt{VERB}}
\newcommand{\deter}{\texttt{det}}
\newcommand{\adpmod}{\texttt{adpmod}}
\newcommand{\adpobj}{\texttt{adpobj}}
\newcommand{\advmod}{\texttt{advmod}}
\newcommand{\appos}{\texttt{appos}}
\newcommand{\cc}{\texttt{cc}}
\newcommand{\ccomp}{\texttt{ccomp}}
\newcommand{\conj}{\texttt{conj}}
\newcommand{\dobj}{\texttt{dobj}}
\newcommand{\marker}{\texttt{mark}}
\newcommand{\nmod}{\texttt{nmod}}
\newcommand{\nsubj}{\texttt{nsubj}}
\newcommand{\p}{\texttt{p}}
\newcommand{\prt}{\texttt{prt}}
\newcommand{\rcmod}{\texttt{rcmod}}
\newcommand{\rel}{\texttt{rel}}

\noindent {\bf 6.} As per the CoNLL-X format, the first token in a sentence is numbered 1. 

\noindent {\bf 7.} In general, break on whitespace.

\noindent {\bf 8.} Separate out commas, colons, semicolons, exclamation marks, full stops and question marks.

\noindent {\bf 9.} If a parenthesis introduces ordinary text then separate it out.

\noindent {\bf 10.} If on the other hand it indicates a list item or an abbreviation, for example ``(3)'' or ``(iii)'' then keep it together.

\noindent {\bf 11.} Keep hyphens together with the text they join up.

\noindent {\bf 12.} Separate out \textit{m'} and \textit{d'} but retain the apostrophe. Join up \textit{dh'}.

\noindent {\bf 13.} Also keep the apostrophe in \textit{`S} and \textit{a'} (this is important!).

\section{Annotation rules word for word}

\subsection{\textit{A}\label{subsect:a}}

\begin{table}
\begin{tabular}{l l l l l}
Function &  Rule & Category & Rule \\\hline 
Agreement &  --- & \tt S[n]\textbackslash N/S[n]/N/N & \bf 102 \\
Embedder &  \bf **59 & \tt S[em]/S[dcl] & \bf **51 \\
Interrogative &  \textbf{17} (subsect.~\ref{subsect:verbs}) & \tt S[q]/S[dep] & \textbf{22} (subsect.~\ref{subsect:verbs}) \\
Numerical &  --- & **** &---\\
Possessive &  \bf 92 & \tt N/N & \bf 95 \\
Relativizer &  \textbf{55} (subsect.~\ref{subsect:is}) & \tt (N\textbackslash N)/S[dcl] & \bf 76 \\
Vocative &  \bf 107 & **** & ---\\
In \textit{a cheile} &  \bf 87 & \tt N/N \rm ex.gd40& \bf 86 \\
\hline
\end{tabular}
\caption{\textit{A}.\label{table:a}}
\end{table}

The various functions of \textit{a} are described in Table~\ref{table:a}.
It is possible that we can merge the embedder into the relativizer, which would involve altering the types in Table~\ref{tab:interrog}.

\subsection{\textit{A'}\label{subsect:ag}}

See subsection~\ref{subsect:determiners} for the article and subsection~\ref{subsect:prepositions} for the form of \textit{ag} before a verbal noun.

\subsection{Adjectives\label{subsect:adjectives}}
\newcommand{\NbN}{\texttt{N\textbackslash N}}
\newcommand{\SsS}{\texttt{S/S}}
\newcommand{\SbNsSbN}{\texttt{S\textbackslash N/S\textbackslash N}}

\subsubsection*{Categorial}

\noindent{\bf 28.} Most attributive adjectives are annotated as \texttt{S[adj]/N} but then type-changed on parsing to \texttt{N\textbackslash N}.
\textit{\`Oran m\`or}.

\noindent{\bf 29.} Prenominal attributive adjectives are \texttt{N/N}.
Examples (possibly not exhaustive) \textit{ath, corra, deagh, dearbh, droch, fior, s\'ar} and \textit{seann}.

\noindent{\bf 30.} Predicative adjectives are \texttt{S[adj]/N}.

\noindent{\bf 63.} A predicative phrase such as \textit{na bu dl\`uithe} will be \texttt{S[adj]/N} as a whole.
\textit{bu}, being a form of \textit{is} will be \texttt{S[dcl]/S[adj]/N/N}. This means that \textit{na} is \texttt{(S[adj]/N)/S[dcl]}.

\noindent{\bf 84.} Adjectives that subcategorize with a PP, for example \textit{measail}, take type \texttt{S[adj]/N/PP[*]}.
\textit{(Consider whether these type-change to \texttt{N\textbackslash N/PP[*]})}.

\subsection{Adverbs\label{subsect:adverbs}}

\subsubsection*{Categorial}

\newcommand{\SsNbSsN}{\texttt{S/N\textbackslash S/N}}

\noindent{\bf **104.} Treat manner adverbs that have the form of a PP as PPs.

\noindent{\bf 113.} Corollary: this means that \textit{ais} in \textit{air ais}
is a noun.

\noindent{\bf **105.} (This turns out to be a rule that I have been applying implicitly.) 
Adverbs are \SsNbSsN.
They will type-change to \SsS\ if they are preposed.



\subsection{\textit{An}}

For \textit{am} and \textit{an} the articles see subsection~\ref{subsect:determiners}.
For \textit{am} and \textit{an} the interrogative particles see subsections~\ref{subsect:bi} and \ref{subsect:is}.
For \textit{am} and \textit{an} the possessive pronouns see subsection~\ref{subsect:pronouns}.
For \textit{am} and \textit{an} the prepositions see subsection~\ref{subsect:prepositions}.

\subsection{\textit{Ann}}

\noindent{\bf **98.} Treat as a preposition or PP, even where it looks adverbial.

\subsection{\textit{Bi}\label{subsect:bi}}



\subsubsection*{Categorial}

\noindent{\bf 58.} \textit{Bi} and variants are \begin{itemize}
\item \texttt{S[*]/PP[*]/N}: \textit{Tha Mairead anns a' chidsin.} (non-verbal use of prepositions)
\item \texttt{S[*]/S[adj]/N/N}: \textit{Tha mi sgith}.
\item \texttt{S[*]/S[asp]/N/N}: \textit{Tha mi a' seinn}.
\item \texttt{S[*]/PP[ri]/PP[aig]}: obligation. See also rule 83 under prepositions.
\end{itemize}

Where the experiencer is marked in a PP modifier of the experience, use one of the following patterns:
\begin{itemize}
\item \texttt{S[*]/S[gu]/N}: \textit{Bha an t-eagal orm gum biodh daoine a' magadh orm.} (gd7)
\item \texttt{S[*]/S[wq]/N}: \textit{Tha fios agam a-nise carson a tha falt fada aig March.} (gd6)
\item \texttt{S[*]/S[q]/N}: knowing, say, whether
\end{itemize}

However, where the experience is a PP modifier of the experiencer, as in sentence gd20: \textit{Bha na poilis an d\`ochas gum biodh MacCaluim air fhaighinn ciontach de mhurt.}, treat it as in case \bf 58a\rm.



\subsection{Coordinators\label{subsect:coordinators}}

These are \textit{agus}, \textit{is}, \textit{'s}, \textit{ach} and \textit{no}.
For other conjunctions see subsection \ref{subsect:subordinators}.

\subsubsection*{Categorial}

\noindent{\bf 69.} Mark conjunctions as \texttt{conj}. 

\subsection{Determiners\label{subsect:determiners}}

\subsubsection*{Categorial}

\noindent{\bf 32.} Articles (see rule {\bf 18}) and \textit{gach} are \texttt{N/N}.

\noindent{\bf ***38.} Determiners of attributive nouns are \texttt{(N\textbackslash N)/(N\textbackslash N)}.

\subsection{\textit{Do, d', dh'}\label{subsect:do}}

For \textit{do} the preposition, see subsection~\ref{subsect:prepositions},
for \textit{do} (\textit{d'}) the possessive pronoun, see subsection~\ref{subsect:pronouns}, and
for \textit{do} (\textit{dh'}) the past tense marker, see subsection~\ref{subsect:verbs}.

\subsection{\textit{Fh\`ein}\label{subsect:fhein}}

\subsubsection*{Categorial}
\noindent{\bf 44.} \textit{Fh\`ein} is \texttt{N\textbackslash N}.
See the type-changing rules for what the parser should do if it follows a fused preposition--possessive pronoun.

\subsection{\textit{Gach}\label{subsect:gach}}

See subsection~\ref{subsect:determiners}.

\subsection{\textit{Gu}\label{subsect:gu}}

See subsection~\ref{subsect:prepositions}.

\subsection{Interrogatives\label{subsect:interrogatives}}

\noindent{\bf 51 \rm and \bf 52.} See Table~\ref{tab:interrog} for non-polar interrogatives.
This table also deals with their relative counterparts which behave somewhat like subordinators.
\begin{center}
\begin{table}
\begin{tabular}{l l l l}
Question form & POS & Category \\ \hline
\textit{c\`aite} & \ADV &  \texttt{S[wq]/S[q]} \\ 
\textit{c\`o} & \texttt{NOUN} & \texttt{S[wq]/S[em]}, \texttt{S[wq]/S[dcl]} \\
\textit{d\`e} & \texttt{NOUN} & \texttt{S[wq]/S[em]}, \texttt{S[wq]/S[dcl]} \\
\textit{carson} & \ADV & \texttt{S[wq]/S[em]} \\
\textit{cuine} & \ADV & \texttt{S[wq]/S[em]} \\
\textit{ciamar} & \ADV & \texttt{S[wq]/S[em]}
\smallskip\\\hline
Relative form & POS & Category \\\hline
\textit{far} & \PRT & \SbNsSbN\texttt{/S[em]}\\
\textit{c\`o} & \PRT & \texttt{N/S[em]}, \texttt{N/S[dcl]}\\
\textit{na} & \PRT & \texttt{N/S[em]}, \texttt{N/S[dcl]}\\
\textit{airson 's} & \PRT & **** \\
\textit{nuair} & \PRT & \SbNsSbN\texttt{/S[em]}\\
\textit{mar} & \PRT &\texttt{N/S[em]} \\
\hline
\end{tabular}
\caption{Analysis based on examples (37--42) in Lamb (2003).
Note anomalous category of \textit{c\`aite} and also that \textit{c\`o} and \textit{d\`e} need not have the embedder \textit{a} immediately after them.
**** indicates the cases that I haven't worked out yet.
\label{tab:interrog}}
\end{table}
\end{center}
\subsection{\textit{Is} (copula, or assertive)\label{subsect:is}}

\texttt{VERB} in Universal POS tagset.



\subsubsection*{Categorial}
\noindent{\bf **74.} 
The cases as shown in Rule \textbf{89}: nominal experience, adjectival experience, nominal target and verbal target, expand as follows:
\begin{itemize}
\item \texttt{S[*]/N/N} (nominal experience and target)
\item \texttt{S[*]/S[adj]/N/N} (adjectival experience, nominal target)
\item \texttt{S[*]/N/S[n]/N} (nominal experience and verbal target)
\item \texttt{S[*]/S[adj]/N/S[n]/N} (adjectival experience, verbal target)
\end{itemize}

\noindent {\bf 75.} For \textit{is} constructions equating two things, use \texttt{S[*]/N$_2$/N$_1$} (omitting the indices), where NP$_1$ is \textit{e} and NP$_2$ is, say, \textit{tidsear a th'annam}. 

\subsection{\textit{Nach}\label{subsect:nach}}

\texttt{PRT} in Universal POS tagset.

\noindent {\bf 62.} In the case of, say, \textit{M\`airi nach maireann} (Mary who is no longer with us), \textit{maireann} is an \texttt{xcomp} of \textit{M\`airi} and \textit{nach} is the \texttt{prt} of \textit{maireann}.


\subsection{Nouns\label{subsect:nouns}}

\texttt{NOUN} in Universal POS tagset.

\subsubsection*{Categorial}

Initially we won't distinguish between \texttt{N} and \texttt{NP}.

\noindent{\bf 31.} Heads of NPs are \texttt{N}.

\noindent{\bf ***33.} Nouns used attributively, in the genitive, are \texttt{N\textbackslash N} because they follow the head.

\noindent{\bf 34,35.} Verbal nouns in a small clause, are \texttt{S[n]/N/N} if transitive and \texttt{S[n]/N} if intransitive.

\subsection{Numbers\label{subsect:numbers}}

Use \texttt{NUM} from the Universal tagset for simple numbers, \textit{aon}, \textit{ochd}, \textit{ceud}.

\subsubsection*{Categorial}

\noindent{\bf 109.} Attributive numbers are \texttt{N/N}.

\subsection{Prepositions and aspect markers\label{subsect:prepositions}}

In the Universal POS tagset, prepositions are \texttt{ADP} except when they are acting as aspect markers, when they are \texttt{PRT}.



\subsubsection*{Categorial}
\noindent {\bf 24.} Prepositions fused with pronouns take type \texttt{PP[*]} where * is the unfused form of the preposition.
Hence \textit{air} (meaning ``on him'') and \textit{oirre} are both \texttt{PP[air]}.

\noindent {\bf 25.} Unfused prepositions governing a non-verbal noun are \texttt{PP[*]/N}.

\noindent {\bf 26.} \textit{a'}, \textit{ag}, \textit{ann}, \textit{air}, \textit{gu} and \textit{ri} when governing a verbal noun are \texttt{S[asp]/S[n]}.

\noindent {\bf 27.} Prepositions fused with articles are \texttt{PP[*]/N} as in rule {\bf 25}.

\noindent {\bf 66.} Remember that \textit{an} can be a preposition, as in for example \textit{an d\`ochas}.

\noindent {\bf **83.} \textit{Ri} in expressions of obligation and \textit{airson} in order to do something are \texttt{PP[*]/S[n]/N}. Examples: gd28 and gd35.

\subsection{Pronouns\label{subsect:pronouns}}

These are \texttt{PRON} in the Universal POS tagset. In many cases in these guidelines pronouns are included among the nouns.



\subsubsection*{Categorial}

\noindent {\bf 94.} Pronouns in the nominative are \texttt{N}.

\noindent {\bf 95.} Possessive pronouns are \texttt{N/N}.

\noindent {\bf 96.} Fused preposition--possessive pronouns are \texttt{PP[*]/N}.
If they are aspect markers then they are \texttt{(S[asp]/N)/(S[n]/N/N)}.

\subsection{Punctuation\label{subsect:punctuation}}

Punctuation marks are \texttt{.} in the Universal POS tagset.


\subsubsection*{Categorial}

\noindent{\bf 39.} Type full stops as \texttt{.}, commas as \texttt{,} and so forth.

\noindent {\bf 81.} Again, treat commas and colons as coordination in lists and run-on sentences.

\subsection{Subordinators\label{subsect:subordinators}}

\subsubsection*{Categorial}
\noindent{\bf **48.} Because \textit{gu} and variants are arguments of the verbs \textit{bi} and \textit{is} where the other argument is an NP headed by a psych noun like \textit{fios} or \textit{eagal}, treat the \textit{gu} clause as \texttt{S[gu]}.
Hence \textit{gu} itself will be \texttt{S[gu]/S[dep]}.
\textit{Cf.} rule \textbf{110}. 


\subsection{Verbs, verbal nouns and complements\label{subsect:verbs}}

Verbs are \texttt{VERB} in the Universal POS tagset.
Verbal nouns are also \texttt{VERB} in this annotation scheme.




\subsubsection*{Categorial}
\noindent {\bf 21.} There are five sorts of sentence-level clauses, \texttt{S[dcl]} (declarative), \texttt{S[q]} (interrogative), \texttt{S[neq]} (negative), \texttt{S[negq]} (negative interrogative) and \texttt{S[wq]} for wh-questions.

\noindent{\bf **50.} One level down, we have \texttt{S[em]} for embedded declaratives.
\textit{Ciamar a tha thu?}

\noindent{\bf **110.} Where a clause headed by \textit{gu} is an argument to a verb, as in, say, \textit{Tha fhios gu...}, the whole clause is \texttt{S[gu]}.
Annotate as \texttt{S[gu]} anyway.
It will type-change to \texttt{N\textbackslash N} or \texttt{S\textbackslash S} otherwise.

\noindent{\bf **101.} There are (at least) two possible approaches to infinitival clauses, for example \textit{E\`oin a ph\`osadh}.
One is to treat the agreement particle \textit{a} as an infinitivizer, to say that \textit{air} as an aspect marker expects \texttt{S[inf]/N} to its right, and for intransitive small clauses, like \textit{air falbh}, to type-change \texttt{S[n]/N} to \texttt{S[inf]/N}.
What we do here is just to treat infinitival clauses exactly the same as verbal noun clauses, hence \texttt{S[n]/N}.
See rule {\bf 102} in subsection~\ref{subsect:a} for the implications for the agreement particle.

\noindent {\bf 22.} Preverbal particles marking negative, interrogative and negative-interrogative clauses are \texttt{S[*]/S[dep]}.

\noindent {\bf 23.} The past-tense marker takes type \texttt{S[dep]/S[dep]} if following a preverbal particle: \textit{Cha do rinn .}
If starting a sentence the past-tense marker takes type \texttt{S[dcl]/S[dcl]}: \textit{Dh' fhaicinn .}

\noindent{\bf 40.} Intransitive clauses take the type \texttt{S[*]/N}; transitive \texttt{S[*]/N/N}.

\noindent{\bf 45.} Infinitive clauses (\textit{a} followed by a lenited verbal noun and preceded by an object) as a whole are \texttt{S[n]/N}.

\noindent{\bf ***47.} \textit{Rach} introducing a passive is \texttt{S[*]/S[n]/N/N}.
Bizarre example in gd38, \textit{cf.}~example (98) in Lamb.
See (96) and (97) in Lamb for more normal examples.

\noindent {\bf 58.} For \textit{bi} see subsection~\ref{subsect:bi}.

\noindent {\bf 74} and {\bf 75.} For \textit{is} see subsection \ref{subsect:is}.


\section{Type-changing rules\label{section:typechanging}}

These are all either feature-changing rules or turn an argument into an adjunct.

\begin{itemize}
\item \texttt{S[adj]/N} $\rightarrow$ \texttt{N\textbackslash N}
\item \texttt{PP} $\rightarrow$ \texttt{N\textbackslash N}
\item \texttt{PP} $\rightarrow$ \texttt{S\textbackslash S}
\item \texttt{PP} $\rightarrow$ \texttt{(S/N)\textbackslash(S/N)}.
Special rule to deal with unstressed prepositions; see example (132c) in Lamb: \textit{Chunnaic mi san eaglais an-diugh e}.
\item \texttt{PP} $\rightarrow$ \texttt{S/S} (preposed PPs)
\item \SbNsSbN $\rightarrow$ \SsS\ (preposed adverbs)
\item \texttt{S[asp]/N} $\rightarrow$ \texttt{N\textbackslash N}.
Example (gd33): \textit{``se\`oladairean a' seinn''}.
\item \texttt{S[n]/N} $\rightarrow$ \texttt{S[asp]/N}.
Example (gd34): \textit{``Chan eil mi str\`i''}.
\item \texttt{S[gu]} $\rightarrow$ \NbN
\end{itemize}

This is a rule of the type \texttt{X\textbackslash X} $\rightarrow$ \texttt{Y\textbackslash Y}.

\begin{itemize}
\item \texttt{N\textbackslash N} $\rightarrow$ \texttt{PP\textbackslash PP}.
Special rule to deal with \textit{fh\`ein} following a fused preposition--possessive pronoun.
\end{itemize}

\end{document}